Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-02-06 analysis/paper_draft3.Rmd
updated introduction, fixed @Bennett reference

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal bias. Current methods for detecting bias make implicit, but unacknowledged, assumptions about how to incorporate historical strike data into the bias detection process. Here, we show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of alpha influences the estimation and detection of bias in the face of varying degrees of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application (R Shiny application) to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with real, historical strike data from a convenience sample of cases from one court.

@@

A _Batson_ challenge typically proceeds in a four-step sequence (Figure \@ref(fig:batsonworkflowfig)). First, the attorney must decide whether to challenge the opposing attorney's use of one or more peremptory strikes as based on illegal bias. Second, if challenged, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient initial inference of illegal bias. Third, the striking attorney must proffer permissible reasons for their challenged strikes. Fourth, the challenger must discredit those proffered reasons and ultimately persuade the trial judge that it is more likely than not that the striking attorney acted with illegal bias.

For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes [Wash. General Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1); Conn. Superior Court Rule 5-12(d)]. Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

@@

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). Ignoring historical strike data increases the risk of underpowered or inflated estimates of attorney strike bias, because of the low number of strikes per trial (typically 6, 10, or 15 strikes per party).

Relying on historical strike data, however, requires an approach to  dealing with incompatibility between current and historical trail data. Such incompatibility might arise due to missingness in the historical data or to the possibility that attorney's strike bias depends on certain trial-level characteristics that vary across trials (e.g., defendant race, charge severity). As a result, incompatibility might cause an attorney's bias parameter based on strike data for an attorney in the current trial to differ from such values based on historical strike data for that attorney.

Simply pooling the the strike data from current and past trials, and analyzing such data together, ignores such incompatibility, which can lead to inflated type I error and reduced power [@Bennett2020]. To increase power, one may use other methods, such as stratifying strike data by trial and applying the Cochran-Mantel-Haenszel (CMH) test [@Gastwirth2014, 295-297]. Such approaches, however, do not directly address incompatibility between current and historical strike data. For example, applying the CMH test requires first selecting only historical strike data from past trials that are "similar" to the current trial [@Gastwirth2014, 289, 297]. Assessing such similarity implies weighting the available historical strike data with respect to some attorney-bias-related set of trial-level characteristics that are present or absent in the current trial.

@@

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows us to explicitly and transparently adjust for our assumptions about the degree of incompatibility between current and historical trials. In this approach, we use two steps. In step one, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. We use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. We use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter. In step two, we estimate the posterior distribution of the bias parameter from the current case, using the posterior distribution developed with the historical information and the power prior.

The use of the power prior in this context allows lawyers to explicitly control the assumed degree of incompatibility between the current and historical trial data. One approach is to fix the power prior parameter $\alpha$ and assess the sensitivity of results to different values [@Carvalho2021, 5252; @Ibrahim2015, 3734]. Specifically, lower values (closer to 0) of $\alpha$ should be selected if incompatibility is expected to be high. Higher values (closer to 1) of the power prior parameter should be selected if incompatibility is expected to be low. Accordingly, the lawyer bringing the _Batson_ challenge, after choosing the value of $\alpha$, must be able to defend that choice as appropriately corresponding to the likely degree of incompatibility between the current and historical strike data in the particular case. Since this overall procedure is to estimate bias to support the bringing of a _Batson_ challenge, the senstivity of bias detection to the value of $\alpha$ is used to evaluate sensitivity of the estimation of bias, not to indicate whether a _Batson_ challenge should succeed.

Here, we demonstrate this approach with simulation studies where we vary the degree of incompatibility between current and historical trial data, and test our ability to detect bias using different values of alpha. Then, we present a software prototype that encodes the same approach with actual historical strike data. With that data, the prototype shows how attorneys and other can this approach to help them decide to argue, and ultimately prove, illegal bias in peremptory-challenge use in real time during jury selection.

@@

```{r inpts}
df <- data.frame(round = c(1:10),
                 num_cog = c(5,5,4,3,2,2,2,2,2,2),
                 total = c(21,20,19,18,17,16,15,14,13,12),
                 cog = c(0,1,1,0,0,1,1,0,0,0),
                 party = rep(c("PP","PD"),5))

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal bias. Current methods for detecting bias make implicit, but unacknowledged, assumptions about how to incorporate historical strike data into the bias detection process. Here, we show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences the bias detection (whether the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

A _Batson_ challenge typically proceeds in a four-step sequence. First, the attorney must decide whether to challenge the opposing attorney's use of one or more peremptory strikes as based on illegal bias. Second, if challenged, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient initial inference of illegal bias. Third, the striking attorney must proffer permissible reasons for their challenged strikes. Fourth, the challenger must discredit those proffered reasons and ultimately persuade the trial judge that it is more likely than not that the striking attorney acted with illegal bias.

For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes [Wash. General Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1); Conn. Superior Court Rule 5-12(d)]. Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@Williams2017].

@@

Without historical strike data, there is a higher risk of underpowered or inflated estimates of attorney strike bias, because of the low number of strikes per trial (typically 6, 10, or 15 strikes per party, depending on the jurisdiction) [@nscs-jury2018]. Yet, completely pooling available current and past strike data ignores possible incompatibility between current and historical strike data. Such incompatibility might arise due to missingness in the historical data or because an attorney's strike bias depends on certain trial-level characteristics that vary across trials (e.g., defendant race, charge severity).

As a result, incompatibility might cause an attorney's bias parameter based on strike data for an attorney in the current trial to differ from such values based on historical strike data for that attorney. This can lead to inflated type I error and reduced power [@Bennett2021]. To increase power, one may use other methods, such as stratifying strike data by trial and applying the Cochran-Mantel-Haenszel (CMH) test [@Gastwirth2014, p. 295-297]. Such approaches, however, often adjust for incompatibility implicitly. For example, applying the CMH test requires first selecting only historical strike data from the subset of past trials that are "similar" in its bias-salient characteristics to the current trial [@Gastwirth2014, p. 289, 297], a step to reduce incompatibility due to variation in those salient characteristics.

@@

```
In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows for transparently adjusting for the assumed degree of incompatibility between current and historical trials. For this approach, we proceed in three steps.  First, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Second, we use an attorney's strike data from past trials, generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the power prior [@Chen2000; @Ibrahim2015], which raises the likelihood of the historical strike data to a fixed power ($\alpha_{0}$) between zero and one, to weight the historical strike data in accord with expected compatibility between the historical trials and the current trial. Zero denotes no weight to the historical data (complete incompatibility) and one denotes equal weight (complete compatibility). Third, we use this posterior distribution as an informative prior for the bias parameter in the current trial.

Here, we demonstrate this approach with simulations in which we vary the degree of incompatibility between current and historical trial data, and test how well we can "detect" attorney strike bias (i.e., whether the credible interval for the bias parameter excludes zero), given different values of the power prior's alpha parameter. Then, we present a software prototype that encodes the same approach with actual historical strike data. With that data, the prototype shows how attorneys and others can use this approach to detect strike bias in real time during jury selection.

To be sure, neither this nor any other statistical approach can automatically detect _Batson_ violations, because _Batson_ law permits any evidence of strike bias, not just the attorney's strikes in past cases. And it requires the trial judge to weigh all such evidence to decide whether strike bias likely exists (the _Batson_ violation). Still, _before_ accounting for other relevant evidence of strike bias, our approach can help attorneys "detect" strike bias, given current and historical strike data, the model, and a fixed value for the power prior's $\alpha$ parameter. After testing for sensitivity of bias detection to different fixed $\alpha$ values [@Carvalho2021, p. 5252; @Ibrahim2015, p. 3734], attorneys can combine the resulting inference with other evidence to decide whether to argue _Batson_; to prove the _Batson_ prima facie case; and to persuade the trial judge to find  a _Batson_ violation (Figure \@ref(fig:batsonworkflowfig)).

@@

```{r inpts}

df <- data.frame(round = c(1:10),
                 num_cog = c(5,5,4,3,2,2,2,2,2,2),
                 total = c(21,20,19,18,17,16,15,14,13,12),
                 cog = c(0,1,1,0,0,1,1,0,0,0),
                 party = rep(c("PP","PD"),5))
